---
title: "Patch-based segmentation and regression with rcissus"
author: "Brian B. Avants"
date: "`r Sys.Date()`"
bibliography: REFERENCES.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Patch-based segmentation and regression with rcissus}
  %\usepackage[UTF-8]{inputenc}
---

```{r global options, include=FALSE}
library( rcissus )
```


# RCISSUS

After Narcissus, who sought his reflection.

A fast approach to patch-based deep learning segmentation or image translation
(mapping intensities between imaging modalities).

In examples below, the epochs are limited to allow for the examples to run quickly.
One should increase the epochs to get better results.

# Regression

An example predicting raw intensity from gradient and laplacian images.

```{r regression}
if ( ! exists( "mth" ) ) {
  mth = 'kerasdnn' # may require more tuning than h2o but can yield better results
  mth = 'h2o'
  }
nBases = 5
nEpochs = 150
scaleData = 10000
# step 1 - collect data
library( rcissus )
popfns = getANTsRData('show')[1:5]
testfn = getANTsRData('show')[6]
# below, we use image lists but one can alternatively replaces lists
# with vectors of filenames - just do so consistently
popGT = list( )  # ground truth
popGR = list( )  # gradient
popLA = list( )  # laplacian
masks = list( )  # sample masks
nsam = 5000 # samples
myPR = 3
mc = FALSE
for ( i in 1:length( popfns ) ) {
  popGT[[ i ]] = antsImageRead( getANTsRData( popfns[ i ] ) )
  popGR[[ i ]] = iMath( popGT[[ i ]], "Grad", 1 )
  popLA[[ i ]] = iMath( popGT[[ i ]], "Laplacian", 1 )
  masks[[ i ]] = randomMask( thresholdImage( popGT[[ i ]], 1 , 255  ) , nsam )
  }
trnBas = rcBasis( lappend( popGR, popLA  ), patchRadius = myPR, meanCenter = mc )
trnBas$basisMat = trnBas$basisMat[ 1:nBases,  ] # select basis vectors
myseeds = c( 1:length( popGT ) )
trnMat1 = rcTrainingMatrix( popGT, popGR, masks, trnBas, seeds = myseeds, patchRadius = myPR, meanCenter = mc   )
trnMat2 = rcTrainingMatrix( popGT, popLA, masks, trnBas, seeds = myseeds, patchRadius = myPR, meanCenter = mc   )
print( table( trnMat2$y==trnMat1$y ) )

popGTtest = list( )  # ground truth
popGRtest = list( )  # gradient
popLAtest = list( )  # laplacian
maskstest = list( )  # sample masks
for ( i in 1:length( testfn ) ) {
  popGTtest[[ i ]] = antsImageRead( getANTsRData( testfn[ i ] ) )
  popGRtest[[ i ]] = iMath( popGTtest[[ i ]], "Grad", 1 )
  popLAtest[[ i ]] = iMath( popGTtest[[ i ]], "Laplacian", 1 )
  maskstest[[ i ]] = getMask( popGTtest[[ i ]] ) # NOTE: dense prediction!
  }
testMat1 = rcTestingMatrix( popGRtest, maskstest, trnBas, seeds = 1, patchRadius = myPR, meanCenter = mc  )
testMat2 = rcTestingMatrix( popLAtest, maskstest, trnBas, seeds = 1, patchRadius = myPR, meanCenter = mc  )


traindf = data.frame( trnMat1$x, trnMat2$x, trnMat1$position ) / 100
testdf = data.frame( testMat1$x, testMat2$x, testMat1$position ) / 100
# if h2o works on your machine, use deep learning
trn = rcTrain( trnMat1$y, traindf, mdlMethod = mth, epochs = nEpochs )
prd = rcPredict( trn, testdf, mdlMethod = mth )
mm = makeImage( maskstest[[1]], as.numeric( prd[,1] ) )
plot( mm )

```

# Segmentation

The same approach can be used for segmentation.

```{r segmentation,eval=TRUE}
# step 1 - collect data
nBases = 5
# below, we use image lists but one can alternatively replaces lists
# with vectors of filenames - just do so consistently
popGT = list( )  # ground truth
popGR = list( )  # gradient
masks = list( )  # sample masks
nsam = 5000 # samples
myPR = 4
for ( i in 1:length( popfns ) ) {
  popGT[[ i ]] = antsImageRead( getANTsRData( popfns[ i ] ) )
  popGR[[ i ]] = thresholdImage( popGT[[ i ]], "Otsu", 3 )
  masks[[ i ]] = randomMask( thresholdImage( popGT[[ i ]], 1, 255  ) , nsam )
  }
trnBas = rcBasis( popGT, patchRadius = myPR )
trnBas$basisMat = trnBas$basisMat[  1:nBases,  ] # select basis vectors
myseeds = c( 1:length( popGT ) )
trnMat1 = rcTrainingMatrix( popGR, popGT, masks, trnBas, seeds = myseeds, patchRadius = myPR  )

# step 2 - build similar data for prediction - but use a dense mask
popGTtest = list( )  # ground truth
popGRtest = list( )  # gradient
maskstest = list( )  # sample masks
for ( i in 1:length( testfn ) ) {
  popGTtest[[ i ]] = antsImageRead( getANTsRData( testfn[ i ] ) )
  popGRtest[[ i ]] = thresholdImage( popGTtest[[ i ]], "Otsu", 3 )
  maskstest[[ i ]] = getMask( popGTtest[[ i ]] ) # NOTE: dense prediction!
  }
testMat1 = rcTestingMatrix( popGTtest, maskstest, trnBas, seeds = 1, patchRadius = myPR )

# step 3 - now implement the training and testing
traindf = data.frame( trnMat1$x, position=trnMat1$position ) / scaleData
testdf = data.frame( testMat1$x, position=testMat1$position ) / scaleData

# if h2o works on your machine, use deep learning
trn = rcTrain( trnMat1$y, traindf, classification = TRUE, mdlMethod = mth, epochs = nEpochs )
prd = rcPredict( trn, testdf, classification = TRUE, mdlMethod = mth )
prdnumerics = as.numeric(prd$predict) - min( as.numeric(prd$predict) )
mm = makeImage( maskstest[[1]], prdnumerics  )
plot( popGTtest[[ 1 ]], mm, window.overay = range( mm ), alpha = 0.6 )
sel = maskstest[[ 1 ]] == 1
gtvals = popGRtest[[ 1 ]][ sel ]
print( table( gtvals == prdnumerics  )  )
plot( popGTtest[[ 1 ]], popGRtest[[ 1 ]], window.overay = range( mm ), alpha = 0.6 )
mm = makeImage( maskstest[[1]], prd[ , "class_2" ]  )
plot( mm  )
```


# Further optimizations

Note: one may want to customize the deep learning components used here, which
are simply the defaults provided in h2o.  One can inspect the code in order to
get started on this, as well as the h2o documentation.  In order to optimize
performance for a specific problem, one will need to define a validation objective
and take advantage of the many parameter search strategies available.




# References
